---
title: "Final_Markdown"
author: "Jonathon Sun"
date: "11/20/2021"
output: html_document
---

Link to canva presentation: https://www.canva.com/design/DAEx6RRkde0/share/preview?token=u2xD2CeqLfZqDMSWH58_Dw&role=EDITOR&utm_content=DAEx6RRkde0&utm_campaign=designshare&utm_medium=link&utm_source=sharebutton 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, vroom, DataExplorer, lubridate, sf, tidycensus, tigris, hms, riem, slider, scales, caret,RcppEigen, ckanr, FNN, ggcorplot, grid, gridExtra, gtools, httr, kableExtra, modelsummary,tidycensus, spdep, stargazer, tmap, viridis, zipcodeR, DT, dplyr, scales, ggthemes, plotly, gganimate)

library(ggcorrplot)

#RcppEigen https://rdrr.io/cran/RcppEigen/ runs fastLm()

options(scipen = 100)

root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette_9_colors <- c("#FF2AD4","#E53AD8","#CC4ADC","#996AE5","#7F7BE9",
                      "#668BED","#33ABF6","#19BBFA","#00CCFF")
palette_3_colors <- c("#FF2AD4","#7F7BE9","#00CCFF")
palette_2_colors <- c("#FF2AD4", "#00CCFF")
palette_1_colors <- c("#00CCFF")
```

# Case
Project option 4 - Forecast Metro train delays in and around NYC: An amazing new dataset has popped up on Kaggle recently that list origin/destinations delays for Amtrak and NJ Transit trains. Can you predict train delays? Consider the time frame that it would be useful to have such predictions. Predicting 5 minutes out is not going to be as useful as 2-3 hours out. Consider training on a month and predicting for the next week or two. Consider time/space (train line, county etc.) cross validation. Many app use cases here.


```{r}
import <- paste0("Amtrak/",list.files("Amtrak"))

df <- read.csv(import[1])

for (i in 2:nrow(as_tibble(import))) {
  df <- rbind(df,read.csv(import[i]))
}

df <- df %>%
  filter(type == "NJ Transit") %>%
  mutate(date = ymd(date),
         scheduled_time = ymd_hms(scheduled_time),
         actual_time = ymd_hms(actual_time),
         week = week(date),
         DotW = wday(date),
         Start_time = as_hms(scheduled_time),
         Actual_time = as_hms(actual_time),
         AM_PM = ifelse(am(Actual_time) == TRUE, "AM","PM"),
         Year = year(date),
         train_lag = lag(train_id),
         time_lag_dumb = lag(delay_minutes),
         train_end = train_lag == train_id,
         station_lag = ifelse(train_end == TRUE, time_lag_dumb,NA),
         accumulated_delay = delay_minutes + station_lag,
         quintile_delay = q5(delay_minutes),
         ID = row_number()) %>%
  select(!c("time_lag_dumb","train_end","train_lag")) %>%
  rename(scheduled_datetime = scheduled_time,
         actual_datetime = actual_time)

Analysis_Dates <- seq(min(as.Date(df$date)), max(as.Date(df$date)), by="days")
Analysis_Dates <- as.tibble(Analysis_Dates) %>%
                  mutate(day = row_number()) %>%
                  rename(date = value)

df <- left_join(df,Analysis_Dates, by = "date")


```


# Exploring Data types




# Train Routes
```{r message=FALSE, warning=FALSE, include=FALSE}
# There's no complete set of NJ transit lines anymore

NJTransitLines <- bind_rows(st_read("https://opendata.arcgis.com/datasets/fc24118ac4a0445da5e09f7692075757_13.geojson"),
                            st_read("https://opendata.arcgis.com/datasets/fc24118ac4a0445da5e09f7692075757_9.geojson"),
                            st_read("https://opendata.arcgis.com/datasets/fc24118ac4a0445da5e09f7692075757_3.geojson"),
                            st_read("https://opendata.arcgis.com/datasets/fc24118ac4a0445da5e09f7692075757_4.geojson"),
                            st_read("https://opendata.arcgis.com/datasets/fc24118ac4a0445da5e09f7692075757_1.geojson"),
                            st_read("https://opendata.arcgis.com/datasets/fc24118ac4a0445da5e09f7692075757_5.geojson"),
                            st_read("https://opendata.arcgis.com/datasets/fc24118ac4a0445da5e09f7692075757_0.geojson"),
                            st_read("https://services6.arcgis.com/M0t0HPE53pFK525U/arcgis/rest/services/NJTransit_Rail_Linea/FeatureServer/2/query?outFields=*&where=1%3D1&f=geojson"),
                            st_read("https://services6.arcgis.com/M0t0HPE53pFK525U/arcgis/rest/services/NJTransit_Rail_Linea/FeatureServer/6/query?outFields=*&where=1%3D1&f=geojson"),
                            st_read("https://services6.arcgis.com/M0t0HPE53pFK525U/arcgis/rest/services/NJTransit_Rail_Linea/FeatureServer/7/query?outFields=*&where=1%3D1&f=geojson"),
                            st_read("https://services6.arcgis.com/M0t0HPE53pFK525U/arcgis/rest/services/NJTransit_Rail_Linea/FeatureServer/8/query?outFields=*&where=1%3D1&f=geojson"),
                            st_read("https://services6.arcgis.com/M0t0HPE53pFK525U/arcgis/rest/services/NJTransit_Rail_Linea/FeatureServer/10/query?outFields=*&where=1%3D1&f=geojson"),
                            st_read("https://services6.arcgis.com/M0t0HPE53pFK525U/arcgis/rest/services/NJTransit_Rail_Linea/FeatureServer/11/query?outFields=*&where=1%3D1&f=geojson"),
                            st_read("https://services6.arcgis.com/M0t0HPE53pFK525U/arcgis/rest/services/NJTransit_Rail_Linea/FeatureServer/12/query?outFields=*&where=1%3D1&f=geojson")) %>%
  select(!c("OBJECTID","LINE_CODE"))



NJtransitstations <- st_read("https://opendata.arcgis.com/datasets/4809dada94c542e0beff00600ee930f6_0.geojson") 

NJshp <- tigris::states() %>%
             filter(NAME == "New Jersey")
```


```{r}
DataExplorer::introduce(df)
DataExplorer::plot_intro(df)
DataExplorer::plot_density(df)
DataExplorer::plot_bar(df)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
DataExplorer::plot_histogram(df)
```

Additionally, we've included a correlation table to show how the variables we made are or are not correlated.


```{r echo=FALSE, message=FALSE, warning=FALSE}
DataExplorer::plot_correlation(df)
```

# Attaching Geography
```{r}
#df %>%
#  filter(line %in% c("Amtrak","AMTRAK","AMTRAK REGIONAL")) %>%
#  View()

#f <- df %>%
#      mutate(merge = case_when(from == "New York Penn Station" & to == "New York Penn Station" & line == "Amtrak" ~ "Winner",
#                               from == "Philadelphia" & to == "New York Penn Station" & line == "Amtrak" ~ "Something",
#                               from == "New York Penn Station" & to == "Philadelphia" & line == "Amtrak" ~ "Apple"))

# Attaching weather -----------------------------------
weather.Panel <- 
  riem_measures(station = "EWR", date_start = (as.Date(min(df$date))-5), date_end = (as.Date(max(df$date))+5)) %>%
  dplyr::select(valid, tmpf, p01i, sknt)%>%
  replace(is.na(.), 0) %>%
    mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
    mutate(week = week(interval60),
           dotw = wday(interval60, label=TRUE)) %>%
    group_by(interval60) %>%
    summarize(Temperature = max(tmpf),
              Precipitation = sum(p01i),
              Wind_Speed = max(sknt)) %>%
    mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))

df <- left_join(df %>% 
                        mutate(interval60 = floor_date(actual_datetime, unit = "hour")), weather.Panel, by = "interval60")



df <- df %>%
              mutate(clean = ifelse(is.na(Start_time) == TRUE, "remove","keep")) %>%
              filter(clean == "keep") %>%
              select(!clean) %>% 
              mutate(LINE_NAME = case_when(line == "Northeast Corrdr" ~ "Northeast Corridor",
                                        line == "No Jersey Coast" ~ "NJ Coast Line",
                                        line == "Main Line" ~ "Main Line",
                                        line == "Morristown Line" ~ "Morristown Line",
                                        line == "Gladstone Branch" ~ "Gladstone Branch",
                                        line == "Raritan Valley" ~ "Raritan Valley Line",
                                        line == "Bergn Co. Line" ~ "Bergen County Line",
                                        line == "Atl. City LIne" ~ "Atlnatic City Line",
                                        line == "Montclair-Boonton" ~ "Montclair-Boonton Line",
                                        line == "Princeton Shuttle" ~ "Princeton Dinky",
                                        line == "Pascack Valley" ~ "Pascack Valley Line")) 

df_shp <- left_join(df, NJTransitLines, by = "LINE_NAME") %>%
            st_as_sf()
```

```{r}
NJtransitstations[149,2] <- "Middletown NY"
NJtransitstations$STATION_ID <- trimws(NJtransitstations$STATION_ID)

Station_clean <- df_shp %>%
                   st_drop_geometry() %>%
                   select(from) %>%
                   arrange(from) %>%
                   unique() %>%
                   mutate(Spelled_correctly = ifelse(from %in% NJtransitstations$STATION_ID, TRUE, FALSE)) %>%
                   filter(Spelled_correctly == FALSE)


f <- c("Anderson Street-Hackensack",
       "Atlantic City",
  "Bay Street-Montclair",
  "Broadway",
  "Essex Street-Hackensack",
  "Glen Rock-Boro Hall",
  "Glen Rock-Main",
  "Hoboken Terminal",
  "Middletown",
  "Montclair St Univ",
  "Mt Arlington",
  "Mountain Station",
  "Pennsauken Transit Center",
  "30th Street Station",
  "Princeton Jct.",
  "Radburn",
  "Ramsey",
  "Rte 17 Ramsey",
  "Secaucus Junction Upper Level",
  "Secaucus Junction Lower Level",
  "Secaucus Junction Upper Level",
  "Teterboro-Williams Ave",
  "Watsessing",
  "Wayne Route 23 Transit Center",
  "Wood-Ridge")

Station_clean <- bind_cols(Station_clean,f) %>%
                  rename(clean_name = ...3)

for(i in 1:nrow(Station_clean)) {
df_shp$from <- str_replace(df_shp$from, Station_clean$from[i],Station_clean$clean_name[i])
df_shp$to <- str_replace(df_shp$to, Station_clean$from[i],Station_clean$clean_name[i])
}

```

# Exploratory Data

```{r}
numericVars <- df_shp %>%
                st_drop_geometry() %>%
                na.omit() %>%
                select(where(is.numeric)) %>%
                select(!c("from_id","to_id","DotW","station_lag","stop_sequence","Year","week"))

ggcorrplot(
  round(cor(numericVars), 1), 
  p.mat = cor_pmat(numericVars),
  colors = c("#25CB10", "white", "#FA7800"),
  type="lower",
  insig = "blank") +  
    labs(title = "Correlation across numeric variables")


```

## Categorical

```{r}
Categorical <- df_shp %>%
                st_drop_geometry() %>%
                na.omit() %>%
                select(where(is.character),delay_minutes)

colnames(Categorical)

cat <- Categorical %>%
        select(colnames(Categorical)[2], delay_minutes) %>%
        mutate(Variable = colnames(Categorical[2])) %>%
        dplyr::rename("Value" = 1)

for(i in 2:(nrow(as.tibble(colnames(Categorical)))-1)) {
t <- Categorical %>%
        select(colnames(Categorical)[i], delay_minutes) %>%
        mutate(Variable = colnames(Categorical[i])) %>%
        dplyr::rename("Value" = 1)

cat <- bind_rows(cat,t)
}
```


```{r}
ggplot(data = cat %>%
         filter(!Variable %in% c("from","to","train_id","type")),
       aes(x = Value,  y = delay_minutes)) +
  geom_bar(stat = "summary", fun.y = "mean") +
  scale_y_continuous(labels = comma) +
  facet_wrap(~Variable, scales = "free") +
  plotTheme() +
  labs(x = "Categorical Variables",
       y = "Delay in Minutes",
       title = "Cateogrical Variables Average Delay")

#ggsave("Categorical_Variables.jpg", width = 16, height = 9, unit = "in", dpi = 300)
```

## station delays

```{r fig.height=12}
ggplot(data = cat %>%
         filter(Variable %in% c("from","to")) %>%
         arrange(Value) %>%
         mutate(Value = as.factor(Value)),
       aes(x = Value,  y = delay_minutes, fill = Variable)) +
  geom_bar(position = "dodge2", stat = "summary", fun.y = "mean") +
  scale_y_continuous(labels = comma) +
  plotTheme() +
  labs(x = "Categorical Variables",
       y = "Delay in Minutes",
       title = "Cateogrical Variables Average Delay") +
  coord_flip()

#ggsave("Station_Delay.jpg", width = 16*1.5, height = 9*1.5, unit = "in", dpi = 300)
```

## station delays by geography

```{r}
A <- df_shp %>%
  st_drop_geometry() %>%
  group_by(from) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  mutate(Bin_mean_delay = cut_interval(mean_delay, n = 20, dig.lab = 10))

A <- length(unique(A$Bin_mean_delay))

cc <- scales::seq_gradient_pal("Red","Grey")(seq(0,1,length.out=A))

df_shp %>%
  st_drop_geometry() %>%
  group_by(from) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  mutate(Bin_mean_delay = cut_interval(mean_delay, n = 20, dig.lab = 10),
         Bin_mean_delay = str_remove(Bin_mean_delay,"."),
         Bin_mean_delay = str_sub(Bin_mean_delay,1,nchar(Bin_mean_delay)-1)) %>%
  separate(Bin_mean_delay, into = c("Start","End"), sep = ",") %>%
  mutate(Start = as.numeric(Start)) %>%
  arrange(Start) %>%
  mutate(Start = as.character(Start),
         Interval = paste0(Start," - ",End),
         Interval = forcats::as_factor(Interval)) %>%
  rename(STATION_ID = from) %>%
  left_join(NJtransitstations, by = "STATION_ID") %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(data = NJTransitLines) +
  geom_sf(aes(color = Interval)) +
  geom_sf(data = NJshp, fill = "transparent") +
  scale_color_manual(values = cc) +
  labs(title = "Delay in minutes going from a Station") +
  mapTheme()

#ggsave("Delay_from_station.jpg", width = 16, height = 9, unit = "in", dpi = 300)
```

## Line delays by geography

```{r}

A <- df_shp %>%
  st_drop_geometry() %>%
  group_by(to) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  mutate(Bin_mean_delay = cut_interval(mean_delay, n = 20, dig.lab = 10))

A <- length(unique(A$Bin_mean_delay))

cc <- scales::seq_gradient_pal("Red","Grey")(seq(0,1,length.out=A))

df_shp %>%
  st_drop_geometry() %>%
  group_by(to) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  mutate(Bin_mean_delay = cut_interval(mean_delay, n = 20, dig.lab = 10),
         Bin_mean_delay = str_remove(Bin_mean_delay,"."),
         Bin_mean_delay = str_sub(Bin_mean_delay,1,nchar(Bin_mean_delay)-1)) %>%
  separate(Bin_mean_delay, into = c("Start","End"), sep = ",") %>%
  mutate(Start = as.numeric(Start)) %>%
  arrange(Start) %>%
  mutate(Start = as.character(Start),
         Interval = paste0(Start," - ",End),
         Interval = forcats::as_factor(Interval)) %>%
  rename(STATION_ID = to) %>%
  left_join(NJtransitstations, by = "STATION_ID") %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(data = NJTransitLines) +
  geom_sf(aes(color = Interval)) +
  geom_sf(data = NJshp, fill = "transparent") +
  scale_color_manual(values = cc) +
  labs(title = "Delay in minutes going to a Station") +
  mapTheme()

ggsave("Delay_to_station.jpg", width = 16, height = 9, unit = "in", dpi = 300)
```
## Train delays

```{r eval=FALSE, fig.height=, include=FALSE}
ggplot(data = cat %>%
         filter(Variable %in% c("train_id")),
       aes(x = Value,  y = delay_minutes)) +
  geom_bar(position = "dodge2", stat = "summary", fun.y = "mean") +
  scale_y_continuous(labels = comma) +
  plotTheme() +
  labs(x = "Categorical Variables",
       y = "Delay in Minutes",
       title = "Cateogrical Variables Average Delay") +
  coord_flip()

```


## Delays over Time

```{r}
df_shp %>%
  st_drop_geometry() %>%
  group_by(date,LINE_NAME) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  ggplot(aes(x = date, y = mean_delay, color = LINE_NAME)) +
  geom_smooth(se = FALSE) +
  geom_vline(xintercept = as.numeric(ymd(c("2020-01-20"))), linetype = "dashed") +
  labs(title = "Average Delay Over Time",
       subtitle = "By Line",
       y = "Average Delay",
       x = "Date") +
  plotTheme()

#ggsave("smooth_line_delay.jpg", width = 16, height = 9, unit = "in", dpi = 300)
```



```{r}
covid_day <- 691

animate(
  df_shp %>%
  st_drop_geometry() %>%
  group_by(day,LINE_NAME) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  na.omit() %>%
  ggplot(aes(x = day, y = mean_delay, color = LINE_NAME)) +
  geom_line(se = FALSE) +
  geom_vline(xintercept = covid_day, linetype = "dashed") +
  labs(title = "Average Delay Over Time",
       subtitle = "By Line",
       y = "Average Delay",
       x = "Day") +
  plotTheme() +
  theme(legend.position = "hide") +
  facet_wrap(~LINE_NAME)+
  transition_reveal(day), 
  width = 800, height = 800)

#anim_save(filename = "LINE_NAME.gif",animation = last_animation())
```



```{r}
df_shp %>%
  st_drop_geometry() %>%
  group_by(date,status) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  ggplot(aes(x = date, y = mean_delay, color = status)) +
  geom_smooth(se = FALSE) +
  geom_vline(xintercept = as.numeric(ymd(c("2020-01-20"))), linetype = "dashed") +
  labs(title = "Average Delay Over Time",
       subtitle = "By Line",
       y = "Average Delay",
       x = "Date") +
  plotTheme()

factor_order <- c(1:26)

df_shp %>%
  st_drop_geometry() %>%
  group_by(date,stop_sequence) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  mutate(stop_sequence = factor(stop_sequence, levels = factor_order)) %>% 
  ggplot(aes(x = date, y = mean_delay, color = stop_sequence)) +
  geom_smooth(se = FALSE) +
  geom_vline(xintercept = as.numeric(ymd(c("2020-01-20"))), linetype = "dashed") +
  labs(title = "Average Delay Over Time",
       subtitle = "By Line",
       y = "Average Delay",
       x = "Date") +
  plotTheme()

#ggsave("smooth_Stop_delay.jpg", width = 16, height = 9, unit = "in", dpi = 300)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
animate(
  df_shp %>%
  st_drop_geometry() %>%
  group_by(day,stop_sequence) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  mutate(stop_sequence = factor(stop_sequence, levels = factor_order)) %>% 
  ggplot(aes(x = day, y = mean_delay, color = stop_sequence)) +
  geom_smooth(se = FALSE) +
  geom_vline(xintercept = covid_day, linetype = "dashed") +
  labs(title = "Average Delay Over Time",
       subtitle = "By Line",
       y = "Average Delay",
       x = "Date") +
  plotTheme() +
  facet_wrap(~stop_sequence) +
  theme(legend.position = "hide",
        axis.ticks.x = element_blank()) +
  transition_reveal(day), 
  width = 800, height = 800)

#anim_save(filename = "stop_sequence.gif",animation = last_animation())
```

```{r}
f <- df_shp %>%
  st_drop_geometry() %>%
  group_by(date, LINE_NAME) %>%
  summarize(mean_delay = mean(delay_minutes)) %>%
  ggplot(aes(x = LINE_NAME, y = mean_delay)) +
  geom_boxplot()

ggplotly(f)
```

# Model building


```{r}
set.seed(111)

#sample_size = round(nrow(df_shp %>%
#                           st_drop_geometry())*.6)
#trainIndex <- sample(seq_len(nrow(df_shp %>%
#                           st_drop_geometry())), size = sample_size)

#Train <- lmdf[trainIndex,]
#Valid <- lmdf[-trainIndex,]



remove <- c("from_id","to_id","type","interval60","line","date","scheduled_datetime","actual_datetime","train_id")

Sample <- df %>%
          select(!remove)
Sample <- Sample[sample(nrow(Sample), size = nrow(Sample) *.5), ]

Test <- df %>%
          select(!remove) %>%
          filter(!ID %in% Sample$ID) %>%
          filter(from %in% Sample$from) %>%
          filter(to %in% Sample$to) %>%
          filter(LINE_NAME %in% Sample$LINE_NAME) %>%
          na.omit() 

#to test use slice()

Sample <- Sample %>%
            select(!ID)

Test <- Test %>%
          select(!ID)
gc()

#reg.training <- fastLm(delay_minutes ~ ., data = Sample)

reg.training <- lm(delay_minutes ~ ., data = Sample)

#summary(reg.training) Our R squared is 1. Definitely an overfit model. We'll see how well it predicts though. 

f <- predict(reg.training, Test)

Test <- bind_cols(Test, as.tibble(f)) %>%
          rename(Delay.Predict = value) %>%
          mutate(Delay.Error = Delay.Predict - delay_minutes,
               Delay.AbsError = abs(Delay.Predict - delay_minutes),
               Delay.APE = (abs(Delay.Predict - delay_minutes)) / Delay.Predict)


mae <- mean(Test$Delay.AbsError, na.rm=TRUE)
print(mae)
mape <- mean(Test$Delay.APE, na.rm=TRUE)
print(mape)


# Table of regression outputs
modelsummary(reg.training, output = "kableExtra", stars = TRUE, gof_omit = 'IC|Log|AIC|BIC')
```

# Cross Validation

```{r Cross_Validation}
fitControl <- trainControl(method = "cv", number = 5)
set.seed(825)

reg.cv <- 
  train(delay_minutes ~ ., data = Sample, 
     method = "lm", trControl = fitControl, na.action = na.pass)

reg.cv
```

## Average Predicted Delay by Train Route

```{r}


Test %>%
  group_by(LINE_NAME) %>%
  summarize(Mean_Delay = mean(Delay.Predict)) %>%
  left_join(NJTransitLines, by = "LINE_NAME") %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(color = Mean_Delay)) +
  geom_sf(data = NJshp, fill = "transparent")


```

## Average Predicted Delay From Station 
```{r}
Test %>%
  group_by(from) %>%
  summarize(Mean_Delay = mean(Delay.Predict)) %>%
  rename(STATION_ID = from) %>%
  left_join(NJtransitstations %>%
              select(STATION_ID,LATITUDE,LONGITUDE,geometry), by = "STATION_ID") %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(color = Mean_Delay)) +
  geom_sf(data = NJshp, fill = "transparent") +
  labs(title = "from station average of predicted delay") +
  mapTheme()

```



## Average Predicated Delay To Station

```{r}

Test %>%
  group_by(to) %>%
  summarize(Mean_Delay = mean(Delay.Predict)) %>%
  rename(STATION_ID = to) %>%
  left_join(NJtransitstations %>%
              select(STATION_ID,LATITUDE,LONGITUDE,geometry), by = "STATION_ID") %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(color = Mean_Delay)) +
  geom_sf(data = NJshp, fill = "transparent") +
  labs(title = "to station average of predicted delay") +
  mapTheme()
```


# Appendix


```{r}
df_clean %>%
  group_by(train_id, Year) %>%
  summarize(mean_delay = mean(delay_minutes),
            sd_delay = sd(delay_minutes),
            mean_temperature = mean(Temperature),
            sd_temperature = sd(Temperature),
            mean_precipitation = mean(Precipitation),
            sd_precipitation = sd(Precipitation))

f <- df_clean %>%
  filter(is.na(Temperature))

NA_weather_dates <- unique(f$interval60)

```

```{r}
Amtrak_stops <- bind_rows(
                  df %>%
                    filter(type == "Amtrak") %>%
                    group_by(from) %>%
                    summarize(total = n()) %>%
                    rename(location = from),
                  df %>%
                    filter(type == "Amtrak") %>%
                    group_by(to) %>%
                    summarize(total = n()) %>%
                    rename(location = to)) %>%
                  arrange(location) %>%
                  group_by(location) %>%
                  summarize(total = sum(total))

NJTransit_stops <- bind_rows(
                      df %>%
                        filter(type == "NJ Transit") %>%
                        group_by(from) %>%
                        summarize(total = n()) %>%
                        rename(location = from),
                      df %>%
                        filter(type == "NJ Transit") %>%
                        group_by(to) %>%
                        summarize(total = n()) %>%
                        rename(location = to)) %>%
                      arrange(location) %>%
                      group_by(location) %>%
                      summarize(total = sum(total))

AllStops <- bind_rows(Amtrak_stops, NJTransit_stops) %>%
              group_by(location) %>%
              summarise(total = sum(total))

```

## Amtrak Data

```{r eval=FALSE, include=FALSE}
#Amtrak_routes <- st_read("https://opendata.arcgis.com/datasets/baa5a6c4d4ae4034850e99aaca38cfbb_0.geojson") 
#Amtrak_stations <- st_read("https://opendata.arcgis.com/datasets/4cf728602fa3428ba0a08d30efbb5f45_0.geojson")
# In case amtrak links don't work -------------------------------------

Amtrak_routes <- st_read(".//AmtrakRoutes//amtrakroutes.shp") 

Amtrak_stations <- st_read(".//AmtrakStations//amtrakstations.shp")

#st_write(NJTransitLines,"NJTransitLineShapefile/NJTransitlines.shp")
#st_write(Amtrak_stations,"AmtrakStations/Amtrakstations.shp")
#st_write(Amtrak_routes,"AmtrakRoutes/Amtrakroutes.shp")

# -----------------------------------------
sts <- c("NJ","NY","PA")

combined <- rbind_tigris(
  lapply(sts, function(x) {
    tracts(x, cb = TRUE)
  })
) %>%
  st_transform(st_crs(Amtrak_stations))
  

#AmtrakLinesDf <- st_intersection(combined, Amtrak_routes) %>%
#  group_by(NAME.1) %>%
#  summarize(total = n())

#AmtrakLinesDf %>%
#  select()

#ggplot() +
#  geom_sf(data = combined) +
#  geom_sf(data = AmtrakLinesDf)

#Amtrak_stations %>%
#  separate(StationNam, into = c("County", "State"), sep = ",", remove = FALSE) %>%
#  mutate(State = trimws(State)) %>%
#  filter(State %in% c("NY","NJ","PA")) %>%
#  View()





```


```{r cleaning_Stations}

NJtransitstations %>%
  filter(grepl("Anderson", STATION_ID))

NJtransitstations %>%
  filter(grepl("Atlantic", STATION_ID))

NJtransitstations %>%
  filter(grepl("Bay", STATION_ID))

NJtransitstations %>%
  filter(grepl("Broad", STATION_ID))

NJtransitstations %>%
  filter(grepl("Essex", STATION_ID))

NJtransitstations %>%
  filter(grepl("Glen Rock", STATION_ID))

NJtransitstations %>%
  filter(grepl("Glen Rock", STATION_ID))

NJtransitstations %>%
  filter(grepl("Hoboken", STATION_ID))

NJtransitstations %>%
  filter(grepl("Middletown", STATION_ID))

NJtransitstations %>%
  filter(grepl("Montclair", STATION_ID))

NJtransitstations %>%
  filter(grepl("Arlington", STATION_ID))

NJtransitstations %>%
  filter(grepl("Mountain", STATION_ID))

NJtransitstations %>%
  filter(grepl("Pennsauken", STATION_ID))

NJtransitstations %>%
  filter(grepl("30th", STATION_ID))

NJtransitstations %>%
  filter(grepl("Princeton", STATION_ID))

NJtransitstations %>%
  filter(grepl("Rad", STATION_ID))

NJtransitstations %>%
  filter(grepl("Ramsey", STATION_ID))

NJtransitstations %>%
  filter(grepl("Secaucus", STATION_ID))

NJtransitstations %>%
  filter(grepl("Teterboro", STATION_ID))

NJtransitstations %>%
  filter(grepl("Watsessing", STATION_ID))

NJtransitstations %>%
  filter(grepl("Wayne", STATION_ID))

NJtransitstations %>%
  filter(grepl("Wood", STATION_ID))


```
